Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-01 appendix.tex
A little more detail and a factor of 2?

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and
$$
\hat{w} \ge 1
$$
is equivalent to
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
$$
a + b + c \ge 0,
$$
which, using
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ yields
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}

@@

\begin{proof}
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
From the previous section, $\hat{w}$ is given by the quadratic formula, though now with $\gamma$ and $\beta$ as functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
Since
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
simplifying \eqref{eq:restricted-bl-result} further yields
\begin{equation}
\label{eq:general-bl-result}
-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
\end{equation}
Given the bounds
$$
\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
$$
and
$$
\beta_{L} \le \beta(\hat{x}, \hat{y}),
$$
if
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
$$
then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
\end{proof}

\subsection*{Empirical validation}

>>>>

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
With this we have,
$$
\hat{w} \ge 1 \iff
|2a + b| \leq \sqrt{b^2 - 4ac} \iff
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
a + b + c \ge 0.
$$
Using
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
%EM Something's not quite right here. If we take the previous comment as truth we should get a
%EM -\gamma^2(1 + \beta) term which is what I get via independent calculation.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}

@@

\begin{proof}
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
From the previous section, $\hat{w}$ is given by the quadratic formula, though now with $\gamma$ and $\beta$ as functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
The same derivation as for \eqref{eq:restricted-bl-result} further yields
\begin{equation}
\label{eq:general-bl-result}
-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
\end{equation}
Given the bounds
$$
\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
$$
and
$$
\beta_{L} \le \beta(\hat{x}, \hat{y}),
$$
if
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
$$
then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
\end{proof}
To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
\subsection*{Empirical validation}

### 2018-01-01 joint_inf.tex
noticed minor error; fix plus add clarification
A little more detail and a factor of 2?

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
$$
\end{restatable}
This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We have shown that the method of joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are in fact quite recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint maximum likelihood approach.

@@

>>>>

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
%EM It's a little surprising that "naked" beta shows up here. While it seems like a correct statement, why can we have beta here to make the plots, while in another term we have to bound it with beta_L? Seems like we'd need a beta_U.
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We have shown that maximum-likelihood joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.

Joint maximum-likelihood inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are in fact quite recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint maximum likelihood approach.

@@

As a final reminder, we note that all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
